Software architectures to integrate workflow engines in science gateways

### 2016-04-25 paper.tex
Updated McGill's affiliation
edits to section 3.1

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration effort, robustness of workflow
execution, extensibility, scalability and other specific features. Criteria
break down to specific metrics where \emph{low value indicates good
  performance}. For each criterion, a total score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that summation
makes sense. The
criteria and metrics are explained hereafter.

@@

@@

\emph{Integration effort} measures the development required to build
the architecture. It is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}),
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g. account creation, software installation, etc) which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for integration effort, which measures the total number of
software pieces to develop (pieces include interactions and
components).

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g. unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc. 
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of
interactions and components that need to be modified when a new
element is added. Modification of a component is required when its
code needs to be updated or recompiled (science gateway or workflow
service), or when a new piece of software has to be installed
(infrastructure only). Modification of an interaction is deemed
necessary when the parameters involved in this interaction are
modified.  Extensibility breaks down in 4 metrics depending on the
type of element that has to be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions that need to be
  invoked to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g. GPUs, clouds), or to enforce execution policies 
  (e.g. constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture.

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the lack of scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature could only be
implemented with a nonsensical amount of effort. Four different
features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when a lot of workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load to several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention, which prevents easy implementations of elastic
  engines. \texttt{S$_1$}=2 when new engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture and \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine.
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.

>>>>

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration effort, robustness of workflow
execution, extensibility, scalability and other specific features. Criteria
break down to specific metrics where \emph{low value indicates better
performance}. For each criterion, a total score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that summation
does not incur in loss of accuracy. The
criteria and metrics are explained hereafter.

@@

\footnotesize

@@

\emph{Integration effort} measures the development required to build
the architecture. It is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for integration effort, which measures the total number of
software pieces to develop (pieces include interactions and
components).

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks) or
workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is determined as the number of software
components and interactions that are specific to workflow execution,
which assumes that complex architectures tend to be prone to failure:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc. 
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are represented in red in
the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$}  are summed to obtain the total score for this
criterion, which measures the total number of software pieces that are
specifically involved in the workflow execution.

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of
interactions and components that need to be modified when a new
element is added. Modification of a component is required when its
code needs to be updated or recompiled (science gateway or workflow
service), or when a new piece of software has to be installed
(infrastructure only). Modification of an interaction is deemed
necessary when the parameters involved in this interaction are
modified.  Extensibility breaks down in 4 metrics depending on the
type of element that has to be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g. GPUs, clouds), or to enforce execution policies 
  (e.g. constrain data to remain in a particular network domain).\note{Rafael}{Adding a new engine type may also allow run in different infrastructures. Maybe it is worth to put a sentence to clarify that the metric account for it separately.}
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture.

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the lack of scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature could only be
implemented with a nonsensical amount of effort. Four different
features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load among several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention (e.g., two-factor authentication), which prevents easy 
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine 
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture, and \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine.\note{Rafael}{$S_2$ would be 2 if the paradigm does not allow to distribute the orchestration? e.g., workflow executions implemented within MPI jobs, in Pegasus PMC and dispel4Py.}
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem, and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.
